---
title: "Code For SSI Project"
output: html_document
date: '2022-05-22'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Datasets: Cleaning and Setup:

Berkeley Earth: Global:

Berkeley Earth: U.S:

Central England Climate Data:

```{r}
setwd(dirname(rstudioapi::getSourceEditorContext()$path))
global <- read.csv(file = '../data/global_temp_cleaned.csv', as.is = TRUE)
us <- read.csv(file = '../data/usa_temp_cleaned.csv', as.is = TRUE)
ts_global = ts(data = global$Global_Avg_1, end = 150, frequency = 12)
ts_us = ts(data = us$USA_Avg_1, start = 1, end = 150, frequency = 12)
ts_global_new = ts(data = global$Global_Avg_1, start = 150, end = 170, frequency = 12)
ts_us_new = ts(data = us$USA_Avg_1, start = 150, end = 170, frequency = 12)
```

Periodogram Analysis:

```{r, fig.width=4, fig.height=3}
library(astsa)
temp = ts(data = global$Global_Avg_1, frequency = 12)
per = mvspec(temp, log="no", main="Periodogram")
n = per$n.used
c(length(temp), n)
m = length(per$freq)
y = cbind(1:m, per$spec, per$freq, 1/per$freq)
colnames(y) = c('j','periodogram','frequency','year/cycle')
y[order(y[,'periodogram'], decreasing=TRUE)[1:10], ]
```
```{r, fig.width=4, fig.height=3}
library(astsa)
temp2 = ts(data = us$USA_Avg_1, start = 1, end = 170, frequency = 12)
per = mvspec(temp2, log="no", main="Periodogram")
n = per$n.used
c(length(temp2), n)
m = length(per$freq)
y = cbind(1:m, per$spec, per$freq, 1/per$freq)
colnames(y) = c('j','periodogram','frequency','year/cycle')
y[order(y[,'periodogram'], decreasing=TRUE)[1:10], ]
```

Interpretation: We see no significant cycles within the parameters of our dataset (the periodogram "tail" value used fell outside our dataset), meaning 
there are no significant periodic trends impacting climate. So we can proceed without worrying about trends.


```{r, fig.width=4, fig.height=3}
library(astsa)
per = mvspec(temp2, log="no", main="Periodogram")
n = per$n.used
c(length(temp2), n)
m = length(per$freq)
y = cbind(1:m, per$spec, per$freq, 1/per$freq)
colnames(y) = c('j','periodogram','frequency','year/cycle')
y[order(y[,'periodogram'], decreasing=TRUE)[1:10], ]
```
Interpretation: We see one significant cycle within the parameters of our dataset, that being seasonal data, meaning 
there are no significant periodic trends impacting climate outisde of this. So we can proceed without worrying about trends in all 3 cases


Moving on, let's check for model appropriate conditions.

### 3.2 Conditions for ARIMA Model Creation
We now consider an ARIMA(p,d,q) model with fractional differences.

$$\phi(B)\ (1-B)^d \ x_t = \theta(B)$$

First, we estimate the parameter $d$.For small $\omega$, we can fit a linear regression model of $\log \hat{f}(\omega) \sim \log(|\omega|)$, and our expected slope would be $\approx -2d$. 

```{r, fig.width=4, fig.height=3}
# plot
per101 = mvspec(ts_global, plot=FALSE, spans=100) # estimate spectrum using L=101

n = per101$n.used
freq = 1:(n/2) / n
plot(log(abs(freq[1:30])), log(per101$spec[1:30]), type="l", lty=1, lwd=2,
     xlab="log (|w|)", ylab="log f(w)", ylim=c(-1.3, 0.5), 
     main="Slope Est")
# fit linear model
coef(lm(log(per101$spec[1:30]) ~ log(abs(freq[1:30]))))
```
```{r}
per101 = mvspec(ts_us, plot=FALSE, spans=100) # estimate spectrum using L=101

n = per101$n.used
freq = 1:(n/2) / n
plot(log(abs(freq[1:30])), log(per101$spec[1:30]), type="l", lty=1, lwd=2,
     xlab="log (|w|)", ylab="log f(w)", ylim=c(-1.3, 0.5), 
     main="Slope Est")
# fit linear model
coef(lm(log(per101$spec[1:30]) ~ log(abs(freq[1:30]))))
```

The fitted linear model results show that the coefficient $-2d \approx -0.057, -0.019$ respectively, which is outside the CI of our stdev of temperature so we can have that $d \approx 0$. 

So, we can reduce the model to an ARMA(p,q) process:

```{r}
glob_ann = aggregate(global$Global_Avg_1, list(global$Year), FUN=mean)
ts_annual = ts(data = glob_ann$x, frequency = 12)
acfs = acf2(ts_annual, max.lag=48, main="ACF/PACF")


  

```

```{r, fig.width=4, fig.height=3}
y1 = diff(ts_global, lag=12)
per1 = mvspec(y1, log="no", ylim=c(0,10))
y2 = diff(ts_us, lag=12)
per2 = mvspec(y2, log="no", ylim=c(0,10))
y3 = diff(ts_global_new, lag=12)
per1 = mvspec(y1, log="no", ylim=c(0,10))
y4 = diff(ts_us_new, lag=12)
per2 = mvspec(y2, log="no", ylim=c(0,10))
```


Now we will fit the ARMA(p,q) model for $$y_t: \ \phi'(B)\ y_t = \theta(B)\ w_t$$


First, we look at the sample ACF and PACF to find the potential ARMA model candidates.

```{r}
acfs = acf2(ts_global, max.lag=120, main="ACF/PACF")
acfs = acf2(ts_us, max.lag=120, main="ACF/PACF")

```


The ACF cuts off after 20 months (about 1.66 years). This indicates we might have an MA(20) process since that's where significance of the ACF begins to trail. The PACF gives evidence of a trailing term for every 3 years, so we might benefit from An AR(3) portion.

So, we try an ARMA(12,20) model

```{r}
mod1 = sarima(y1, p=12, d=0, q=20, details=FALSE)
mod1$ttable
mod2 = sarima(y2, p=12, d=0, q=20, details=FALSE)
mod2$ttable
mod1_new = sarima(y3, p=12, d=0, q=20, details=FALSE)
mod1$ttable
mod2_new = sarima(y4, p=12, d=0, q=20, details=FALSE)
mod2$ttable
predict(mod1,n.ahead=10)

```

All the fitted MA coefficients are significant, though the mean term is not. 

Now let's look at the residuals.

```{r}
resid2 = resid(mod2$fit)
Box.test(resid2)
plot(resid2, xlab="time", ylab="residuals", main="Residuals")
abline(h=0, col=2, lty=2, lwd=2)
acfs = acf2(resid2, max.lag=50, main="Residual ACF / PACF")
```

This time, our Pierce-Box test does not reject $H_0$ that residuals are  independently distributed. 



### 3.4 Trend differences Verification


To examine this conclusion, we use the previous ARMA(1,5) model for difference series $y_t$ fit to the data from 1900 to 2000 (the period where warming ramped up most), and then predict 2001-2019.

```{r}

mod_global = arima(ts_global, order=c(12,0,20))
y_pred_1 = predict(mod_global, n.ahead=12*20)$pred
ts_global_new = ts(data = global$Global_Avg_1, start = 150, end = 170, frequency = 12)
resid_global = sum((ts_global_new - y_pred_1)**2)



```

```{r}

mod_us = arima(ts_us, order=c(12,0,20))
y_pred_2 = predict(mod_us, n.ahead=12*20)$pred
ts_global_new = ts(data = us$USA_Avg_1, start = 150, end = 170, frequency = 12)
resid_us = sum((ts_global_new - y_pred_2)**2)

```




```{r}
temp_trunc_global = window(ts_global)
diffed_global = diff(temp_trunc_global, lag=12)
mod = arima(diffed_global, order=c(24,0,24))
y_pred = predict(mod, n.ahead=12*19)$pred
n = length(temp_trunc_global)
x_pred = y_pred[1:12] + temp_trunc_global[(n-11):n]
for (i in 13:length(y_pred)) {
    x_pred[i] = y_pred[i] + x_pred[i-1]
}
x_pred = ts(x_pred, start=151, end=170, frequency=12)


temp_trunc_loc = window(ts_us)
diffed_us = diff(temp_trunc_loc, lag=12)
mod_us = arima(diffed_us, order=c(12,0,12))
y_pred2 = predict(mod_us, n.ahead=12*19)$pred
n = length(temp_trunc_loc)
x_pred2 = y_pred2[1:12] + temp_trunc_loc[(n-11):n]
for (i in 13:length(y_pred2)) {
    x_pred2[i] = y_pred2[i] + x_pred2[i-1]
}
x_pred2 = ts(x_pred2, start=151, end=170, frequency=12)
x

plot(temp_trunc_global,
     main="global temperature time series model", xlab="year", ylab="temperature", ylim = c(8,12))
plot_x_pred = window(x_pred, 151,170)
points(plot_x_pred, col=4, cex=0.3)
legend("bottomright", legend=c("Data","Predict"), 
        lty=1, col=c(1,2), cex=0.6)


plot(temp_trunc_local,
     main="global temperature time series model", xlab="year", ylab="temperature", ylim = c(8,12))
plot_x_pred2 = window(x_pred2, 151,170)
points(plot_x_pred2, col=4, cex=0.3)
legend("bottomright", legend=c("Data","Predict"), 
        lty=1, col=c(1,2), cex=0.6)
```



```{r, fig.width=6, fig.height=4}
# data plot
par(mfrow=c(1,2))
temp_s = ts(ksmooth(time(temp), temp, "normal", bandwidth=5)$y,
            start=c(1659,2), end=c(2019,4), frequency=4)
plot(temp_s, type="l", xlim=c(1659,2020),
     main="5 yr smoothing temp", xlab="year", ylab="temperature", ylim = c(8,12))
plot_x_pred = window(x_pred, c(2002,1),c(2019,4))
points(plot_x_pred, col=4, cex=0.3)
legend("bottomright", legend=c("Data","Predict"), 
        lty=1, col=c(1,2), cex=0.6)


temp_s = ts(ksmooth(time(temp), temp, "normal", bandwidth=10)$y,
            start=c(1659,2), end=c(2019,4), frequency=4)
plot(temp_s, type="l", xlim=c(1659,2020),
     main="10 yr smoothing temp", xlab="year", ylab="temperature", ylim = c(8,12))
plot_x_pred = window(x_pred, c(2002,1),c(2019,4))
points(plot_x_pred, col=4, cex=0.3)
legend("bottomright", legend=c("Data","Predict"), 
        lty=1, col=c(1,2), cex=0.6)
```






arima(x = ts_global, order = c(12, 0, 13))
arima(x = ts_global, order = c(12, 0, 14))
arima(x = ts_global, order = c(12, 0, 11))
arima(x = ts_global, order = c(12, 0, 10))









